lesson 4
========================================================

### Scatterplots and Perceived Audience Size
Notes:
Scatterplots to check the relationship between continuous variables.
***

### Working directory
```{r workdir}
work_dir <- '/Users/RickyLim/Documents/OnlineLearning/DataAnalysisR/'
```

### Scatterplots
Notes:

```{r Scatterplots}
library(ggplot2)
pf <- read.csv(paste0(work_dir, 'Data/pseudo_facebook.tsv'), sep='\t')
head(pf)
qplot(x=age, y=friend_count, data=pf)
```



***

#### What are some things that you notice right away?
Response:
Most of high friend counts are scattered arount the age below 30.
There are some peculiars observations, where old people (older than 60) have high friend counts.

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x=age,y=friend_count), data=pf) +
        geom_point() + xlim(13,90) 

```

***

### Overplotting
Notes:
age is a continous variable.

```{r Overplotting}
ggplot(aes(x=age,y=friend_count), data=pf) +
        geom_jitter(alpha=1/20) + 
        xlim(13,90) 
# alpha 1/20 means 20 points equals to a single dot.

```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:
- To transform the distance in coordinates, particularly if there is an extreme high value.

```{r Coord_trans()}
ggplot(aes(x=age,y=friend_count), data=pf) +
        geom_jitter(alpha=1/20, position=position_jitter(h=0)) + 
        coord_trans(y="sqrt")


```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
head(pf)
ggplot(aes(x=age, y=friendships_initiated), data=pf) +
    geom_point() 

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
head(pf)
ggplot(aes(x=age, y=friendships_initiated), data=pf) +
    geom_jitter(alpha=1/20, position=position_jitter(h=0))+
    xlim(c(13,90)) +
    coord_trans(y='sqrt')

```

***

### Overplotting and Domain Knowledge
Notes:
common functions in dplyr: filter(), group_by(), mutate(), and arrange().

***

### Conditional Means
Notes:

```{r Conditional Means}
library(dplyr)

# create the dataframe of mean and median of friends counts for each age group.

age_groups <- group_by(pf, age)
head(age_groups)
pf.fc_by_age <- summarise(age_groups, 
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())
head(pf.fc_by_age)
pf.fc_by_age <- arrange(pf.fc_by_age, age)

# alternative code

pf.fc_by_age <- pf %>% 
    group_by(age) %>%
    summarise(friend_count_mean = mean(friend_count),
              friend_count_median = median(friend_count),
              n = n()) %>%
    arrange(age)




```

Create your plot!

```{r Conditional Means Plot}
head(pf.fc_by_age)
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +
    geom_line() +
    xlim(c(13,90))

```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}

p <- ggplot(aes(x=age, y=friendships_initiated), data=pf) +
    coord_cartesian(xlim=c(13,70), ylim=c(0,1000))+
    geom_point(alpha=0.05, position=position_jitter(h=0), color='orange')+
    geom_line(stat='summary', fun.y=mean) +
    geom_line(stat='summary', fun.y=quantile, prob=0.1,
              linetype=2, color='blue') +
    geom_line(stat='summary', fun.y=quantile, prob=0.5,
              linetype=2, color='blue') +
    geom_line(stat='summary', fun.y=quantile, prob=0.9,
              linetype=2, color='blue')
p


```

#### What are some of your observations of the plot?
Response:

90% of the age group of 13-30 are less than 500 friends. 
Whereas the older group are bound to 250 friends.
There is a strange observation for the group of age ca.69 in which it has a spike.
***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

From the scatter plot --> draws the histogram plots that has over and underestimation.

***

### Correlation
Notes:

```{r Correlation}
head(pf)
cor(pf$age, pf$friend_count)
cor.test(pf$age, pf$friend_count, method="pearson")
with(pf, cor.test(age, friend_count, method="pearson"))
with(subset(pf, age <= 70), cor.test(age, friend_count, method="spearmon"))

```

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(x=www_likes_received, y=likes_received), data=pf) +
    geom_point()

```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x=www_likes_received, y=likes_received), data=pf) +
    geom_point() + 
    xlim(0,quantile(pf$www_likes_received, 0.95)) +
    ylim(0, quantile(pf$likes_received, 0.95)) +
    geom_smooth(method='lm', color='red')

cor.test(pf$www_likes_received, pf$likes_received)

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}

```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
#install.packages('alr3')
library(alr3)
data(Mitchell)
tail(Mitchell)
summary(Mitchell$Month)


ggplot(aes(x=Month, y=Temp), data=Mitchell) +
    geom_point() +
    scale_x_discrete(breaks=seq(0,203,12))

cor.test(Mitchell$Temp, Mitchell$Month)

```

### Making Sense of Data
Notes:

```{r Making Sense of Data}


```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
head(pf)
pf$age_with_months <- pf$age + (12- pf$dob_month)/12

```

***

### Age with Months Means

```{r Age with Months Means}
pf.fc_by_age_with_months <- pf %>%
                    group_by(age_with_months) %>%
                    summarise(mean_fc=mean(friend_count),
                                median_fc=median(friend_count),
                                n=n()) %>%
                    arrange(age_with_months)
head(pf.fc_by_age_with_months)



```

Programming Assignment
```{r Programming Assignment}
p1 <- ggplot(aes(x=age, y=friend_count_mean), 
             data=subset(pf.fc_by_age, age<71)) +
        geom_line() +  
        geom_smooth(method='lm', color='red')

p2 <- ggplot(aes(x=age_with_months, y=mean_fc), 
             data=subset(pf.fc_by_age_with_months, age_with_months<71))+
        geom_line() +
        geom_smooth(method='lm', color='red')

# within +/- 2.5 years
p3 <- ggplot(aes(x=round(age/5)*5, y=friend_count),
             data=subset(pf, age < 71)) +
        geom_line(stat='summary', fun.y=mean)

library(gridExtra)
grid.arrange(p1, p2, p3, ncol=1)

```

Notes:

There is a trade-off between the noise and the precision.

Less datapoints with wider bin width means that we will estimate means more precisely (as in the wider bin you have more n data to estimate the mean). But it may miss the important feature between the relationship of age and friend_count.

Sol: to use the stat flexible smoothing function. 

***

### Noise in Conditional Means

```{r Noise in Conditional Means}


```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

- Scatterplot
- jitter (to add noise)
- alpha
- ddplyr (piping, group_by, summarise, arrange)
- conditional means
- Sceptical about the units to be plotted in a scatterplot
- correlation test
- Bivariate [lecture](http://dept.stat.lsa.umich.edu/~kshedden/Courses/Stat401/Notes/401-bivariate-slides.pdf)

***

### ProblemSet

```{r}
data(diamonds)
head(diamonds)
ggplot(aes(x=price, y=x), data=diamonds) +
    geom_point()
with(diamonds, cor.test(price, x))
with(diamonds, cor.test(price, y))
with(diamonds, cor.test(price, z))
summary(diamonds$depth)
p<- ggplot(aes(x=depth, y=price), data=diamonds) +
    geom_point(alpha=1/100)+
    scale_x_continuous(breaks=seq(40,80,2))
p
with(diamonds, cor.test(depth, price))

names(diamonds)
head(diamonds)

p <- ggplot(aes(x=price, y=carat), data=diamonds)+
    geom_point()+
    coord_cartesian(xlim=c(0, quantile(diamonds$carat,0.99)),
                    ylim=c(0, quantile(diamonds$price,0.99)))
p

```
The plot shows exponential relationship between the length and the price of diamonds in the given dataset with some outliers.

```{r priceVol}
dia <- diamonds
dia$vol <- dia$x*dia$y*dia$z
head(dia)

p <- ggplot(aes(x=price, y=vol), data=dia) +
        geom_point() 
p

volZero <- NA
dia$volZero <- ifelse(dia$vol==0, TRUE, FALSE)
dia$volZero <- factor(dia$volZero)
table(dia$volZero)
diaNonZero <- subset(dia, vol != 0 & vol < 800)
with(diaNonZero, cor.test(price, vol))

diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
diamonds_selected <- subset(diamonds, volume != 0 & volume < 800)
p <- ggplot(aes(x=volume, y=price), data=diamonds_selected) +
        geom_point(alpha=1/20)+
        geom_smooth(method='lm', color='red') +
        geom_smooth(method='lm', formula = y ~ poly(x, 2, raw=TRUE), color='blue')+
        geom_smooth(method='lm', formula = y ~ poly(x, 3, raw=TRUE), color='green')+
        coord_cartesian(xlim=c(0,400), ylim=c(0,4000))
p

```

```{r meanPriceClarity}
names(diamonds)
price_clarity <- diamonds %>%
                    group_by(clarity) %>%
                    summarise(mean_price = mean(price),
                              median_price = median(price),
                              min_price = min(price),
                              max_price = max(price),
                              n = n()) 
price_color <- diamonds %>%
                    group_by(color) %>%
                    summarise(mean_price = mean(price),
                              median_price = median(price),
                              min_price = min(price),
                              max_price = max(price),
                              n = n()) 
head(price_clarity)
head(price_color)
p1 <- ggplot(aes(x=clarity, y=mean_price), data=price_clarity) + 
        geom_bar(stat='identity')
p2 <- ggplot(aes(x=color, y=mean_price), data=price_color) + 
        geom_bar(stat='identity')
grid.arrange(p1, p2, ncol=1)

```
There is an exponential relationship between the price and the volume with some outliers.
There's an expensive diamond with a volume near 4000 and a cheaper diamond with a volume near 900.

Histogram vs Barplot, blog [here](http://flowingdata.com/2014/02/27/how-to-read-histograms-and-use-them-in-r/)


```{r knitIt}
library(knitrBootstrap)
knit_bootstrap('lesson4_student.rmd')
sessionInfo()
```
