---
title: "Subsetting"
output: html_document
---

General Quiz

  1. What is the result of subsetting a vector with positive integers, negative integers, a logical vector, or a character vector?
  
    - Subsetting with positive integers will select a specific element of the vector at particular index position. In R the index is 1-based
    
    - Subsetting with negative will select out a specific element of the vector at particular index position
    
    - Subsetting with the logical vector will select specific elements corresponding to the TRUE element in the respective logical vector
    
    - Subsetting with the character vector will select specific elements with names corresponding to the elements in the character vector
    
```{r}
x <- c(1,2,3)
x[1]  # positive integer
x[-1] # negative integer
logical.v <- c(TRUE, FALSE, TRUE)
x[logical.v]

names(x) <- c('a', 'b', 'c')
char.v <- c('b', 'c')
x[char.v]
```

  1. What’s the difference between [, [[, and $ when applied to a list?
  
    - `[` means to subset the element of a list at certain index position. 
      It returns a list.
    
    - `[[` means to get the element of a list at that certain index position.
      It returns the content of a list.
    
    - `$` means to get the element of a list based on its name
  
```{r}
list.x <- list('a', 1, TRUE)
list.x[1]
list.x[[1]]

names(list.x) <- c('a', 'b', 'c')
list.x$c
```
    
  1. When should you use drop = FALSE?
  
    - Note that drop = F in subsetting data.frame array is required to preserve the type. The omission of drop = F is common for programming errors that will cause failure in an unexpected and unclear way.
  
```{r}
mat.x <- matrix(1:4, nrow = 2)
colnames(mat.x) <- c('a', 'b')
dimnames(mat.x)
mat.x
drop(mat.x)
```
    - Important note on subsetting data.frame!
    
```{r}
# There's an important difference if you select a single 
# column: matrix subsetting simplifies by default, list 
# subsetting does not.

df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])

str(df["x"]) # subset a column as a list
str(df[, "x"]) # subset a column as a matrix

```
    
  1. If x is a matrix, what does x[] <- 0 do? How is it different to x <- 0?
  
    - `x[] <- 0` is to fill all the element with 0, whereas, `x <- 0` is to assign the matrix object with value 0
    
  1. How can you use a named vector to relabel categorical variables?
  
```{r}
vector.x <- c('M', 'F')
names(vector.x) <- c('Laki', 'Wanita')

categorical.x <- factor(c('M', 'F', 'M', 'M'), levels = vector.x)
categorical.x
```
    
## Atomic vector

- Sorting vector

```{r}
x <- c(3,4,5,1,2)
x[order(x)]
```

## S3 and S4 objects

S3 object is made up of atomic vector, list, and array

S4 object has two additional subsetting operators, i.e, `@` equivalent to `$` and `slot` equivalent to `[[`

## Exercise

  1. Fix each of the following common data frame subsetting errors:
  
```{r}
mtcars[mtcars$cyl == 4, ]
mtcars[1:4, ]
mtcars[mtcars$cyl <= 5,]
mtcars[mtcars$cyl == c(4,6), ]
```
  
  1. Why does x <- 1:5; x[NA] yield five missing values? (Hint: why is it different from x[NA_real_]?)
  
```{r}
x <- 1:5
x[NA]
x[NA_real_]
```
    
    Because NA is a logical vector of length 1 and in R when NA is used to index a vector of length 5, it will be recycled and returns, therefore, 5 NAs
    Note that a missing value in the index always yields a missing value in the output. wherease NA_real_ is a numeric type of NA and it's not recycled along the vector.
    
  1. What does upper.tri() return? How does subsetting a matrix with it work? Do we need any additional subsetting rules to describe its behaviour?
  
```{r}
x <- outer(1:5, 1:5, FUN = "*")
x[upper.tri(x)]
```
    
    upper.tri is used to subset the upper triangle, upper the diagonal matrix. `upper.tri(x)` returns a logical index in the matrix, and then subset the TRUE positions.
    
  1. Why does mtcars[1:20] return a error? How does it differ from the similar mtcars[1:20, ]?
  
```{r}
mtcars[1:20,]
```
    
`mtcars[1:20]` returns an error as the columns selected on mtcars were undefined.
    
  1. Implement your own function that extracts the diagonal entries from a matrix (it should behave like diag(x) where x is a matrix).
  
```{r}
getDiagonal <- function(x){
  
  #############################################
  # Returns the diagonal values from a matrix #
  #                                           #
  # > x <- matrix(1:4, nrow = 2)              #
  # > getDiagonal(x)                          #
  # > y <- matrix(1:6, nrow = 3)              #
  # > getDiagonal(y)                          #
  # > z <- matrix(1:6, nrow = 2)              #
  # > getDiagonal(z)                          #
  #############################################
  d <- dim(x)
  diag_row <- min(d)
  d_index <- matrix(c(seq(1, diag_row), seq(1, diag_row)), ncol = 2, 
                    nrow = diag_row)
  return(x[d_index])
}
```

## Subsetting operators

The importance different between `$` and `[[` is that `$` performs partial matching.

Note that `x$y is equivalent to x[["y", exact = FALSE]`

```{r}
x <- list(abc = 1)
x$a
#> [1] 1
x[["a"]]
#> NULL
```
  
### Exercises
Given a linear model, e.g., `mod <- lm(mpg ~ wt, data = mtcars)`, extract the residual degrees of freedom. Extract the R squared from the model summary (summary(mod))

```{r}
mod <- lm(mpg ~ wt, data = mtcars)

# extract the residual degrees of freedom
mod$residuals

# extract the R squared from the model summary
summary(mod)$r.squared
```

## Applications

### Lookup tables (character subsetting)

```{r}
x <- c("m", "f", "u", "f", "f", "m", "m")
lookup <- c(m = 'male', f = 'female', u = NA)
lookup[x]
unname(lookup[x])
```

### Matching and merging by hand (integer subsetting)

```{r}
grades <- c(1, 2, 2, 3, 1)
info <- data.frame(
  grade = 3:1,
  desc = c("Excellent", "Good", "Poor"),
  fail = c(F, F, T)
)

# using match
id <- match(grades, info$grade)
info[id, ]

# using rownames
rownames(info) <- info$grade
info[as.character(grades),]
```

### Random samples/bootstrap (integer subsetting)

```{r}
df <- data.frame(x = rep(1:3, each = 2), y = 6:1, z = letters[1:6])
df

# Set seed for reproducibility
set.seed(10)

# Randomly reorder
df[sample(nrow(df)), ]

# Select 3 random rows
df[sample(nrow(df), 3), ]

# Select 6 bootstrap replicates
df[sample(nrow(df), 6, rep = T), ]
```

### Expanding aggregated counts (integer subsetting)

```{r}
df <- data.frame(x = c(2, 4, 1), y = c(9, 11, 6), n = c(3, 5, 1))
rep(1:nrow(df), df$n)
df[rep(1:nrow(df), df$n), ]
```

### Removing columns from data frames (character subsetting)

```{r}
# set to NULL
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
df$z <- NULL

# subset the columns you want
df[c("x","y")]

# use setdiff
df[setdiff(names(df), "z")]
```

### Selecting rows based on a condition (logical subsetting)

```{r}
mtcars[mtcars$gear == 5, ]
mtcars[mtcars$gear == 5 & mtcars$cyl == 4, ]

# De Morgan's laws
# !(X & Y) is the same as !X | !Y
# !(X | Y) is the same as !X & !Y
```

### Boolean algebra vs. sets (logical & integer subsetting)

```{r}
x <- sample(10) < 4
which(x)

unwhich <- function(x, n) {
  out <- rep_len(FALSE, n)
  out[x] <- TRUE
  out
}

x[unwhich(which(x),10)]

```

## Final exercise

* How would you randomly permute the columns of a data frame? (This is an important technique in random forests.) Can you simultaneously permute the rows and columns in one step?

`df[sample(nrow(df)), sample(ncols(df))]`

* How would you select a random sample of m rows from a data frame? What if the sample had to be contiguous (i.e., with an initial row, a final row, and every row in between)?

`df[sample(nrow(df) - m + 1, size = 1) + (0:(m - 1)), ]`

* How could you put the columns in a data frame in alphabetical order?

`df[, order(names(df))]`