---
title: "Introduction_Dplyr"
output: html_document
---

## Libraries

```{r}
library(dplyr)
library(hflights)
```

## `data.frame` to `tbl`
tbl_df is a wrapper for df to print the obj fits into the console window. 
The tbl object inherits data.frame object, therefore, any methods from data.frame is applicable for tbl object

```{r}
hflights <- tbl_df(hflights)
hflights

# get the structure of the data
glimpse(hflights)
class(hflights)
```

## `Lookup vector`

Look up vector is the named atomic vector

```{r}
# convert the flight code into its name
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", "YV" = "Mesa")
hflights$UniqueCarrier <- lut[hflights$UniqueCarrier]
glimpse(hflights)

# Build the lookup table for the cancellation code
lut <- c("A" = "carrier", "B" = "weather", "C" = "FFA", "D" = "security", "E"="not cancelled")

# change the empty code into E code for empty
hflights$CancellationCode[hflights$CancellationCode == ""] <- "E"

# map it
hflights$CancellationCode <- lut[hflights$CancellationCode]

# Inspect the resulting raw values of your variables
glimpse(hflights)
```

## Grammar of dplyr

- `select` (column selection)
- `mutate` (new column creation)
- `filter` (row removal)
- `arrange` (row reordering according to single or multiple variables)
- `summarize` (summary stats)

Common organization of the data is to all observations into rows and variables into columns. This data organization is so called tidy data, referred to `tidyr`

`select` and `mutate` are for variables manipulation

`filter` and `arrange` are for observation manipulaton

`summarize` is for group of observation manipulation

## Question: Do delay tend to shrink or grow during the course of the flight?

```{r}
# 4 variables are used here: ActualElapsedTime, AirTime, ArrDelay, DepDelay

# a new copy of hflights without modification of its original hflights
h1 <- select(hflights, ActualElapsedTime, AirTime, ArrDelay, DepDelay)
h2 <- mutate(h1, loss= ArrDelay - DepDelay)

# loss if positive means that the plane lose time during the flight
# loss if negative means that the plane makes up time and becomes less delay
mean(h2$loss, na.rm = T)
hist(h2$loss)
median(h2$loss, na.rm=T)
```

The negative loss means that delays tend to shrink in the course of the flight.
## Wrapper function inside select

dplyr provides 6 helper functions, each of which only works when used inside select().

  - `starts_with("X")`: every name that starts with "X",
  - `ends_with("X")`: every name that ends with "X",
  - `contains("X")`: every name that contains "X",
  - `matches("X")`: every name that matches "X", which can be a regular expression,
  - `num_range("x", 1:5)`: the variables named x01, x02, x03, x04 and x05,
  - `one_of(x)`: every name that appears in x, which should be a character vector.

```{r}
# Return a tbl containing just ArrDelay and DepDelay
select(hflights, contains("Delay"))
# Return a tbl as described in the second exercise, using both helper functions and variable names
select(hflights, UniqueCarrier, ends_with("Num"), contains("Cancell"))

# Return a tbl as described in the third exercise, using only helper functions.
select(hflights, contains("Time"), contains("Delay"))
```

## Filter


### R logical operator to be used in `filter`

x < y, TRUE if x is less than y
x <= y, TRUE if x is less than or equal to y
x == y, TRUE if x equals y
x != y, TRUE if x does not equal y
x >= y, TRUE if x is greater than or equal to y
x > y, TRUE if x is greater than y
x %in% c(a, b, c), TRUE if x is in the vector c(a, b, c)

Question:How many weekend flights flew a distance of more than 1000 miles but had a total taxiing time below 15 minutes?

```{r}
filter(hflights, DayOfWeek %in% c(6,7) & Distance > 1000 & (TaxiIn + TaxiOut )< 15)
```

## Arrange

By default, arrange() arranges the rows from smallest to largest. Rows with the smallest value of the variable will appear at the top of the data set. You can reverse this behavior with the desc() function. arrange() will reorder the rows from largest to smallest values of a variable if you wrap the variable name in desc() before passing it to arrange().

## Summarise

`summarise` works like `mutate` though, it does not copy dataset but returns the summary rows.

```{r}
# Determine the shortest and longest distance flown and save statistics to min_dist and max_dist resp.
s1 <- summarise(hflights, min_dist = min(Distance), max_dist = max(Distance))
```

### Aggregating functions
`summarise` takes the aggregating functions. Here are the commonly used :

  - `min(x)` - minimum value of vector x.
  - `max(x)` - maximum value of vector x.
  - `mean(x)` - mean value of vector x.
  - `median(x)` - median value of vector x.
  - `quantile(x, p)` - pth quantile of vector x.
  - `sd(x)` - standard deviation of vector x.
  - `var(x)` - variance of vector x.
  - `IQR(x)` - Inter Quartile Range (IQR) of vector x.
  - `diff(range(x))` - total range of vector x.
  
  - `first(x)` - The first element of vector x.
  - `last(x)` - The last element of vector x.
  - `nth(x, n)` - The nth element of vector x.
  - `n()` - The number of rows in the data.frame or group of observations that summarise() describes.
  - `n_distinct(x)` - The number of unique values in vector x.


Next to these dplyr-specific functions, you can also turn a logical test into an aggregating function with `sum()` or `mean()`. A logical test returns a vector of TRUE's and FALSE's. When you apply `sum()` or `mean()` to such a vector, R coerces each TRUE to a 1 and each FALSE to a 0. This allows you to find the total number or proportion of observations that passed the test, respectively.

```{r}
# Calculate the summarizing statistics for flights flown by American Airlines (carrier code "American")
aa <- filter(hflights, UniqueCarrier == "American")
s2 <- summarise(aa, n_flights = n(), n_canc = sum(Cancelled), p_canc = n_canc/n_flights * 100, avg_delay = mean(ArrDelay, na.rm=T))
```

## Piping or Chaining Functions

Piping `%>%`, works as `obj %>% func(x)`, e.g

```{r}
c(1,2,3) %>% sum
```

Question: Drive or fly?

```{r}

#Let's define preferable flights as flights that are 150% faster than driving, i.e. that travel 105 mph or greater in real time. Also, assume that cancelled or diverted flights are less preferable than driving.
# Solve the exercise using a combination of dplyr verbs and %>%
hflights %>%
  select(Dest, Cancelled, Distance, ActualElapsedTime, Diverted) %>%
  mutate(RealTime = ActualElapsedTime + 100, mph = Distance / RealTime * 60) %>%
  filter(mph < 105 | Diverted == 1 | Cancelled == 1) %>%
  summarise(n_non = n(), p_non = n_non/nrow(hflights)*100,
            n_dest = n_distinct(Dest), min_dist = min(Distance), max_dist = max(Distance))
```

## Group-by

```{r}
# hflights is in the workspace as a tbl, with translated carrier names

# Make the calculations to end up with ordered statistics per carrier
hflights %>% 
    group_by(UniqueCarrier) %>%
    summarise(n_flights = n(), 
              n_canc = sum(Cancelled == 1), 
              p_canc = n_canc/n_flights * 100,
              avg_delay = mean(ArrDelay, na.rm=T)) %>%
    arrange(avg_delay, p_canc)

# Answer the question: At which day of the week is total taxiing time highest?
hflights %>% 
    group_by(DayOfWeek) %>%
    summarise(avg_taxi = mean((TaxiIn + TaxiOut), na.rm=T)) %>%
    arrange(desc(avg_taxi))
```

### group-by with mutate rank function

You can also combine `group_by()` with `mutate()`. When you mutate grouped data, `mutate()` will calculate the new variables independently for each group. This is particularly useful when `mutate()` uses the `rank()` function, which calculates within group rankings. `rank()` takes a group of values and calculates the rank of each value within the group, e.g.
`rank(c(21, 22, 24, 23))`
has output
`[1] 1 2 4 3.`

```{r}
#Use group_by() and summarise() to compare the individual carriers. For each carrier, count the total number of flights flown by the carrier (n_flights), the total number of cancelled flights (n_canc), the percentage of cancelled flights (p_canc), and the average arrival delay of the flights whose delay does not equal NA (avg_delay). Once you've calculated these results, arrange() the carriers from low to high by their average arrival delay. Use percentage of flights cancelled to break any ties. Which airline scores best based on these statistics?
#Come up with a way to answer this question: At which day of the week is the average total taxiing time highest? Use group_by(), summarise() and arrange(); you should avoid the use of mutate(). Define the grouped average total taxiing time to be avg_taxi.
# dplyr is loaded, hflights is loaded with fancy carrier names

# Solution to first instruction
hflights %>%
    filter(!is.na(ArrDelay)) %>%
    group_by(UniqueCarrier) %>%
    summarise(p_delay = sum(ArrDelay > 0)/ n()) %>%
    mutate(rank = rank(p_delay)) %>%
    arrange(rank)


# Solution to second instruction
hflights %>%
    filter(!is.na(ArrDelay), ArrDelay > 0) %>%
    group_by(UniqueCarrier) %>%
    summarise(avg = mean(ArrDelay, na.rm = T)) %>%
    mutate(rank = rank(avg)) %>%
    arrange(rank)
```

## Which plane (by tail number) flew out of Houston the most times? How many times? Name the column with this frequency n. Assign the result to adv1. To answer this question precisely, you will have to filter() as a final step to end up with only two numbers in your dataset.

```{r}
adv1 <- hflights %>%
          group_by(TailNum) %>%
          summarise(n = n()) %>%
          filter(n == max(n))

# How many airplanes only flew to one destination from Houston?
adv2 <- hflights %>%
          group_by(TailNum) %>%
          summarise(ndest = n_distinct(Dest)) %>%
          filter(ndest == 1) %>%
          summarise(nplanes = n())


# Find the most visited destination for each carrier 
adv3 <- hflights %>% 
          group_by(UniqueCarrier, Dest) %>%
          summarise(n = n()) %>%
          mutate(rank = rank(desc(n))) %>%
          filter(rank == 1)

# Find the carrier that travels to each destination the most
adv4 <- hflights %>% 
          group_by(Dest, UniqueCarrier) %>%
          summarise(n = n()) %>%
          mutate(rank = rank(desc(n)))%>%
          filter(rank == 1)
```

## `dplyr` on data table

```{r}
library(data.table)
hflights2 <- as.data.table(hflights)

# summarise the number of carrier

s2 <- hflights2 %>%
        summarise(n_carriers = n_distinct(UniqueCarrier))
```

## `dplyr` with `mysql`

```{r}
# set up a src that connects to the mysql database (src_mysql is provided by dplyr)
my_db <- src_mysql(dbname = "dplyr", 
                  host = "dplyr.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                  port = 3306, user = "dplyr", password = "dplyr")
# and reference a table within that src
nycflights <- tbl(my_db, "dplyr")
# nycflights is now available as an R object that references to the remote nycflights table.

# glimpse at nycflights
glimpse(nycflights)

# Calculate the grouped summaries detailed in the instructions.
dbsumm <- nycflights %>%
            group_by(carrier) %>%
            summarise(n_flights = n(),
                      avg_delay = mean(ArrDelay)) %>%
            arrange(avg_delay)
```
