# Lesson 5
========================================================

```{r global_setup, include=FALSE}
opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/', fig.align='center',
               warning=FALSE, message=FALSE)
opts_chunk$set(cache.extra = R.version,cache.path='Cache/')
opts_chunk$set(cache.extra = rand_seed)
```

# Multivariate Data

***

## Moira Perceived Audience Size Colored by Age
Notes:
A scatterplot with more than two variables 
(the addition of another layer, i.e, age coloring)
is difficult to interpret due to overplotting.

***

## Third Qualitative Variable

```{r libraryLoaded, cache=TRUE}
library(ggplot2)
library(dplyr)
library(reshape2)
```

```{r dataset, cache=TRUE}
work_dir <- '/Users/RickyLim/Documents/OnlineLearning/DataAnalysisR/' 
pf <- read.csv(paste0(work_dir,'Data/pseudo_facebook.tsv'), sep='\t')
head(pf)
names(pf)
```

```{r Third Qualitative Variable}
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) +
    geom_boxplot() +
    stat_summary(fun.y = mean, geom='point', shape=4)

ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender)))+
    geom_line(aes(color=gender), stat='summary', fun.y = median)
```

***

## Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries, cache=TRUE}
pf.fc_by_age_gender <- pf %>% 
    filter(!is.na(gender)) %>%
    group_by(age, gender) %>%
    summarise(mean_friend_count = mean(friend_count),
              median_friend_count = median(friend_count),
              n = n()) %>%
    ungroup() %>%
    arrange(age)

head(pf.fc_by_age_gender)
p <- ggplot(aes(x=age, y=median_friend_count), data=pf.fc_by_age_gender) +
        geom_line(aes(color=gender))
p
```


## Reshaping Data
Notes:

Use the reshape2 library, there are two broad formats, mainly the wide and long formats.
More details info can be found [http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/](here)

```{r ratio, cache=TRUE}
# wide format
pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')
head(pf.fc_by_age_gender.wide)

```


***

## Ratio Plot
Notes:

```{r Ratio Plot}
p <- ggplot(aes(x=age, y=female/male), data=pf.fc_by_age_gender.wide)+
        geom_line()+
        geom_hline(yintercept=1, linetype=2, alpha=0.5)
p
```

***

## Third Quantitative Variable (year joined facebook)

```{r Third Quantitative Variable}
head(pf)
pf$year_joined <- with(pf, floor(2014-tenure/365))
```

***

## Cut a Variable
Notes:
- Create the cut variable of year joined facebook.

```{r Cut a Variable}
#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]
summary(pf$year_joined)
pf$year_joined.bucket <- with(pf, cut(pf$year_joined, breaks=c(2004,2009, 2011, 2012, 2014))) 
head(pf)
```

***

## Plotting it All Together
Notes:

```{r Plotting it All Together}
table(pf$year_joined.bucket, useNA='ifany')

p <- ggplot(aes(x=age, y=friend_count), data=subset(pf, !is.na(year_joined.bucket)))+
        geom_line(aes(color=year_joined.bucket), stat='summary', fun.y=median)
p

```

***

## Plot the Grand Mean
Notes:
The group with longer tenure (being in facebook longer than the others, from 2004-2009) has on average higher friend count across ages compared with others. 
This suggests that the longer time people spent on FB, they accumulate more friends across different ages.

```{r Plot the Grand Mean}
p <- ggplot(aes(x=age, y=friend_count), data=subset(pf, !is.na(year_joined.bucket)))+
        geom_line(aes(color=year_joined.bucket), stat='summary', fun.y=mean) +
        geom_line(aes(color='Grand Mean'),stat='summary', fun.y=mean, linetype=2)
p
```

***

## Friending Rate

```{r Friending Rate}
head(pf)
friend_rate <- with(subset(pf, tenure>=1), friend_count/tenure )
head(friend_rate)
summary(friend_rate)

 
with(subset(pf, tenure >= 1), summary(friend_count/tenure))
```

***

## Friendships Initiated
Notes:

What is the median friend rate? `r median(friend_rate)`

What is the maximum friend rate? `r max(friend_rate)`

```{r Friendships Initiated}

p <- ggplot(aes(x=tenure, y=friendships_initiated/tenure), data=subset(pf, tenure >=1))+
        geom_line(aes(color=year_joined.bucket), stat='summary', fun.y=mean)
p


```
Notes:

The most recent tenure typically initiates less friendships.

***

### Bias-Variance Tradeoff Revisited
Notes:
Increasing the bin size will allow more estimates for the mean, therefore the noise (variance) is reduced (high bias but very much less variance).

NOTE: The code changing the binning is substituting x = tenure in the plotting expressions with x = 7 * round(tenure / 7), etc., binning values by the denominator in the round function and then transforming back to the natural scale with the constant in front.

More can be found [http://scott.fortmann-roe.com/docs/BiasVariance.html](here).


```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

p <- ggplot(aes(x = tenure, y = friendships_initiated / tenure),
            data = subset(pf, tenure > 0)) +
        geom_smooth(aes(color = year_joined.bucket))
p
```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:
The prices are discrete variables shown by the histogram. 
When, the bindwith is increased the histogram shows the absence of adjacent spaces among discrete variables, therefore the histogram with binwidth of 10 shows a bias distribution of prices.

```{r Histograms Revisited}
yo <- read.csv(paste0(work_dir, 'Data/yogurt.csv'))
head(yo)
str(yo)

yo$id <- factor(yo$id)
str(yo)
summary(yo$price)

p <- ggplot(aes(x=price), data=yo) +
        geom_histogram()
p

p_10 <- ggplot(aes(x=price), data=yo) +
        geom_histogram(binwidth=10)
p_10

```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
summary(yo)
length(unique(yo$price))
table(yo$price)

yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + mixed.berry)
head(yo)

p <- ggplot(aes(x=all.purchases),data = yo) +
        geom_histogram(binwidth=1, breaks=seq(0,15,1))
p

```

***

### Prices over Time
Notes:

```{r Prices over Time}

p <- ggplot(aes(x=time, y=price), data=yo) +
        geom_jitter(alpha=1/10, shape=21, fill=I('#F79420'))
p

```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

Notes:

x %in% y returns a logical (boolean) vector the same length as x that says whether each entry in x appears in y. That is, for each entry in x, it checks to see whether it is in y.

```{r Looking at Sample of Households}
set.seed(123)
sample.ids <- sample(levels(yo$id), 16)

p <- ggplot(aes(x=time, y=price), 
            data=subset(yo, id %in% sample.ids)) +
        facet_wrap(~id) +
        geom_line()+
        geom_point(aes(size=all.purchases), pch=1)

p

```

***

### The Limits of Cross Sectional Data
Notes:

    - It's a snapshot at a particular time period so no indication of the sequence of events, no causal inference
    - Only a snapshot, the situation may provide differing results if another time-frame had been chosen  
    - Prevalence-incidence bias (neyman bias)


### Scatterplot Matrix
Notes:
- When the data are categorical use boxplot or histogram, otherwise scatterplot is very helpful .

```{r}
library(GGally)
theme_set(theme_minimal(20))

set.seed(12345)
head(pf)
pf_subset <- pf[, c(2:15)]
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])

```

***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table(paste0(work_dir, "Data/nci.tsv"))
colnames(nci) <- c(1:64)
```

```{r}
library(reshape2)
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
        data = nci.long.samp) +
        geom_tile() +
        scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))

p

```


***

# Problem set

## Diamonds

```{r}
head(diamonds)

p <- ggplot(aes(x=price), data=diamonds) +
        facet_wrap(~color) +
        geom_histogram(aes(fill=cut)) +
        scale_fill_brewer(type = 'qual')

p

p <- ggplot(aes(x=table, y=price), data=diamonds)+
        geom_point(aes(color=cut)) +
        scale_color_brewer(type = 'qual')
p

tapply(diamonds$table, diamonds$cut, summary)

diamonds$volume <- with(diamonds, x*y*z)
p <- ggplot(aes(x=volume, y=price), 
            data=subset(diamonds, volume >0)) +
        geom_point(aes(color=clarity)) +
        xlim(0, quantile(diamonds$volume, 0.99))+
        scale_y_log10()+
        scale_color_brewer(type='div')
p

head(diamonds)
p <- ggplot(aes(y=price/carat, x=cut), data=diamonds)+
        geom_jitter(aes(color=color))+
        facet_wrap(~clarity) +
        scale_color_brewer(type = 'div')
p

```

## Facebook

```{r}
# proportions of friends initiated
head(pf)

prop_initiated <- NA
pf$prop_initiated <- with(pf, ifelse(friend_count != 0, 
                          friendships_initiated/friend_count,0))

p <- ggplot(aes(x=tenure, y=prop_initiated),
        data=subset(pf, !is.na(year_joined.bucket)&prop_initiated > 0))+
        geom_line(aes(color=year_joined.bucket), stat='summary',
                  fun.y='mean') +
        geom_smooth(method='auto')

p

p_10 <- ggplot(aes(x=10*round(tenure/10), y=prop_initiated),
        data=subset(pf, !is.na(year_joined.bucket)&prop_initiated > 0))+
        geom_line(aes(color=year_joined.bucket), stat='summary',
                  fun.y='mean') 
p_10
head(pf)

with(subset(pf,!is.na(year_joined.bucket)), 
     by(prop_initiated, year_joined.bucket, mean))
```

## Metainfo

Filename: lesson5_student.rmd

Working directory: `r getwd()`

```{r}
sessionInfo()
```

```{r knitIt, cache=TRUE, results='hide'}
library(knitr)
library(knitrBootstrap)
knit_bootstrap("lesson5_student.rmd")
```
